---
title: "Statistical Inference Assignment - Part One"
author: "Stephanie Lum"
date: "January 31, 2016"
output: pdf_document
---

##Overview
This project investigates exponential distributions simulated in R and compares them to the Central Limit Theorem. The investigation will include the distribution of averages of 40 exponentials and a thousand simulations will be performed.  

This report will show the following simulations and comparisons:  
1. The sample mean vs. the theoretical mean  
2. The sample variance vs. the theoretical variance  
3. The distribution of a large collection of random exponentials vs. the distribution of a large collection of averages of exponentials  


##Simulations
The code below demonstrates the method of finding the mean of 40 exponential distribution values over 1000 simulations.  

```{r}
# Load packages
library(ggplot2)

# Define variables
NumberSim <- 1000  # number of simulations
lambda <- 0.2  # rate parameter
n <- 40 # number of exponentials

# Set the seed to ensure reproducibility
set.seed(8675309)

# Perform 1000 simulations of finding the means of 40 exponential distribution values
SimValues <- matrix(rexp(NumberSim * n, rate = lambda), nrow = NumberSim, ncol = n)
SimMean <- rowMeans(SimValues)
```

A histogram of the means can be plotted.

```{r, echo=FALSE, fig.height=2}
qplot(SimMean,
      geom = "histogram",
      binwidth = 0.1,
      main = "Exponential Mean Distribution, 1000 Simulations",
      xlab = "Means",
      ylab = "Frequency")
      
```

##Sample Mean versus Theoretical Mean
The theoretical mean of the exponential distribution is $\mu = 1/\lambda$ where $\lambda$ is 0.2. 

```{r}
mu <- 1/lambda
mu
```

Therefore, $\mu$ should equal 5.

The sample mean can be seen in the following plot.

```{r, echo=FALSE, fig.height=2}
plot <- qplot(SimMean,
      geom = "histogram",
      binwidth = 0.1,
      main = "Exponential Mean Distribution, 1000 Simulations",
      xlab = "Means",
      ylab = "Frequency")
line <- geom_vline(xintercept = mean(SimMean), col = I("red"))
text <- annotate("text", label = c("Sample Mean:", round(mean(SimMean),3)), x = 6, y = c(60,50))
plot+line+text
```

The sample mean is found to be 5.012, which is very close to the theoretical mean of 5.

##Sample Variance versus Theoretical Variance
The standard deviation of the sample mean of the exponential distribution is $\sigma = \frac{1/\lambda}{\sqrt{n}}$. Therefore, the theoretical variance of the sample mean is $Var = \sigma^2$.

```{r}
# Determine the theoretical variance
sigma <- (1/lambda)/sqrt(n)
Theoretical_Var <- sigma^2
Theoretical_Var
```
The theoretical variance is calculated to be 0.625.

```{r}
# Determine the sample variance
Sample_Var <- round(var(SimMean), 3)
Sample_Var
```
The sample variance is found to be 0.603, which is also very close to the calculated theoretical variance. 

##Distribution
The Central Limit Theorem states that with a large enough sample size, the distribution of averages of iid variables should become that of a standard normal distribution.

A plot of the exponential mean distribution with 1000 simulations is shown below.

```{r, echo=FALSE, fig.height=2}
plot <- ggplot(data.frame(SimMean), aes(SimMean)) + geom_histogram(aes(y=..density..), binwidth = 0.1)
curve <- geom_density(color = "red", size = 1)
plot + curve + labs(title = "Exponential Mean Distribution, 1000 Simulations (Normalized)") + xlab("Means") + ylab("Density")
```

From the plot, it can be seen that the curve of the densities follows the curve of a standard normal distribution.

An additional way to show whether or not this exponential distribution is normal is to test its 95% confidence interval.

```{r, eval=FALSE}
# Calculate the theoretical confidence interval
theoretical_conf_int <- mu + c(-1,1)*1.96*sigma/sqrt(n)

# Calculate the sample confidence interval
sample_conf_int <- mean(SimMean) + c(-1,1)*1.96*sd(SimMean)/sqrt(n)
```

The theoretical 95% confidence interval is calculated to be [4.755, 5.245]. The sample 95% confidence interval is found to be [4.77, 5.25], which is very close to the theoretical interval.

This is evidence that the exponential distribution with a $\lambda$ of 0.2 is approximately normally distributed.